The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
_

\paragraph*{Negative Types.} 
Consider the following algebraic manipulation relating a natural number {{a}}
to itself (ignoring the dotted line for a moment):
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire1.pdf}
}
\end{center}
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~{{a}} (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, this computation `flows along the dotted
line' with the following intermediate steps:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being
  in the left summand of the sum type `{{a+0}}'; we indicate this value
  as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
The example illustrates that the empty type and negative types have a
fascinating computational interpretation that has some flavor of
continuations: negative types denote values that `backtrack to satisfy
dependencies', or in other words act as `debts' that are satisfied by
the backward flow of information.

\paragraph*{Fractional Types.} 
Consider a similar algebraic manipulation involving fractional types.

\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire2.pdf}
}
\end{center}

In the case of negatives, the dotted line indicated the `flow of
control' whereas for fractionals it indicates the `flow of
constraints'. At the heart of logic programming is the idea of
variables that capture constraints and hence it useful to trace this
computation with the analogy to logic variables in mind:

As before, the execution begins at line~(0) with the value {{20}}. At
line~(1) two values, {{20}} and {{()}}, flow forward. One can think of
the value `{{()}}' (of type {{1}}) as `having a credit card': the
`credit card' isn`t money, nor is it debt, but is the option to
generate a `credit-debt' constraint.  At line~(2) we exercise this
option and hence have three values: the initial value {{20}} flowing
from line (1) and two entangled values, {{1/alpha}} and {{alpha}}. The
{{alpha}} and {{1/alpha}} are unspecified values -- i.e. we don't yet
know how much money we need to borrow, but we do know that what is
borrowed must be what is returned. Hence {{alpha}} denotes the
presence of an unknown quantity and dually {{1/alpha}} should be
thought of as the absence of an unknown quantity.

>>>>

\paragraph*{Negative Types.} 
Consider the following algebraic manipulation relating a natural number {{a}}
to itself (ignoring the dotted line for a moment):
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire1.pdf}
}
\end{center}
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~{{a}}, say \$20.00, and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, this computation `flows along the dotted
line' with the following intermediate steps:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being
  in the left summand of the sum type `{{a+0}}'; we indicate this value
  as {{left 20}};
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
The example illustrates that the empty type and negative types have a
fascinating computational interpretation that has some flavor of
continuations: negative types denote values that `backtrack to satisfy
dependencies', or in other words act as `debts' that are satisfied by
the backward flow of information.

\paragraph*{Fractional Types.} 
Consider a similar algebraic manipulation involving fractional types.
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire2.pdf}
}
\end{center}
In the case of negatives, the dotted line indicated the `flow of
control' whereas for fractionals it indicates the `flow of
constraints'. At the heart of logic programming is the idea of
variables that capture constraints. Hence it is useful to trace the
computation corresponding to the algebraic proof above with the
analogy to logic variables in mind.

As before, the execution begins at line~(0) with the value {{20}}. At
line~(1) two values, {{20}} and {{()}}, flow forward. One can think of
the value `{{()}}' (of type {{1}}) as `having a credit card'. The
`credit card' isn't money, nor is it debt, but is the option to
generate a `credit-debt' constraint.  At line~(2) we exercise this
option and hence have three values: the initial value {{20}} flowing
from line (1) and two entangled values, {{1/alpha}} and {{alpha}}. The
{{alpha}} and {{1/alpha}} are unspecified values -- i.e. we don't yet
know how much money we need to borrow, but we do know that what is
borrowed must be what is returned. Hence {{alpha}} denotes the
presence of an unknown quantity and dually {{1/alpha}} should be
thought of as the absence of an unknown quantity.
